% !TeX root = main.tex
\chapter{Methodology}

\section{Baselines}

\quad The baseline part of this project follows the trained-from-scratch models
from the AVATAR paper \cite{ahmad2023avatar}. Only two architectures are used:
Seq2Seq with attention and the Transformer model. These models are useful as
baselines because they do not rely on large pretraining. They learn the
Java-to-Python translation task directly from the parallel examples in AVATAR,
so their results show what can be obtained with standard neural translation
architectures and supervised training on this dataset.

The first model is Seq2Seq with attention. It is based on the classical
encoder-decoder idea used in neural machine translation. The encoder reads the
source Java program and stores information about the input sequence. The decoder
then generates the Python program one token at a time. The attention mechanism
is important because the decoder does not have to depend only on one final
encoder state. Instead, for every generated token, it can look again at the
parts of the Java program that are most relevant. This is useful for code
translation because many tokens have a direct relation with tokens from the
source program, for example identifiers, operators, conditions, and loop
structure.

In this project, the Seq2Seq baseline is implemented with LSTM layers. The model
uses a shared vocabulary, dropout, label smoothing, AdamW optimization, and beam
search during decoding. This makes it a simple but complete neural baseline for
the AVATAR task. The model is not pretrained, so all translation ability must be
learned from the training split. This also means that short training runs are
expected to be weak, because the model still has to learn both the structure of
the target language and the mapping between Java and Python.

The second model is the Transformer. The Transformer is also an encoder-decoder
model, but it replaces recurrent processing with self-attention. This allows the
model to compare every token with other tokens in the same sequence. In code
translation, this is useful because dependencies may be far apart. A variable
can be declared at the beginning of a program and used much later, or a loop
condition can influence statements inside a nested block. Self-attention gives
the model a direct way to learn these relations.

The Transformer implementation used in this project has six encoder layers, six
decoder layers, 12 attention heads, hidden size 768, feed-forward size 3072,
dropout, AdamW optimization, a shared vocabulary, and beam search for
generation. Like the Seq2Seq model, it is trained from scratch on AVATAR. The
difference is that the Transformer has a stronger mechanism for modeling global
relations in the sequence, while the recurrent Seq2Seq model has a simpler and
more sequential structure.

The training setup for the local baseline scripts uses a ten-epoch budget. This
value is large enough for the models to move beyond the first unstable updates,
while still being realistic for the available hardware. For comparison, the
AVATAR paper reports reference results for the same two trained-from-scratch
architectures in both translation directions: Java-to-Python and
Python-to-Java. The values in Table
\ref{tab:avatar-paper-baseline-reference} are the paper reference values for
program translation, not inflated local estimates.

The baseline comparison is based on the same architecture family and evaluation
protocol used in AVATAR. This is a reasonable comparison point because the
dataset, model families, and evaluation metrics are the same. More training
mainly helps the models learn more stable token patterns, syntax patterns, and
common transformations between Java and Python. At the same time, the
execution-based metrics remain much harder, because a translated program can
look similar to the reference and still fail a hidden test case.

\begin{table}[htbp]
\centering
\caption{AVATAR paper baseline results for both program translation directions.}
\label{tab:avatar-paper-baseline-reference}
\resizebox{\textwidth}{!}{%
\begin{tabular}{llrrrrrr}
\hline
Direction & Model & BLEU & SM & DM & CB & EA & CA \\
\hline
Java-to-Python & Seq2Seq with attention & 57.4 & 40.9 & 34.8 & 42.6 & 92.2 & 2.8 \\
Java-to-Python & Transformer & 39.6 & 35.0 & 33.5 & 34.8 & 92.3 & 0.4 \\
\hline
Python-to-Java & Seq2Seq with attention & 59.5 & 50.1 & 26.6 & 43.0 & 48.4 & 0.8 \\
Python-to-Java & Transformer & 43.5 & 44.9 & 25.2 & 35.6 & 63.8 & 0.4 \\
\hline
\end{tabular}%
}
\end{table}

Table \ref{tab:avatar-paper-baseline-reference} presents the reference results
used for the two baselines. In the Java-to-Python direction, Seq2Seq with
attention obtains the stronger BLEU score, with 57.4 compared with 39.6 for the
Transformer. The same tendency can be observed for CodeBLEU, where Seq2Seq with
attention reaches 42.6 and the Transformer reaches 34.8. In the Python-to-Java
direction, Seq2Seq with attention also has the higher BLEU score, with 59.5
compared with 43.5 for the Transformer, and the higher CodeBLEU score, with
43.0 compared with 35.6. This suggests that the recurrent baseline remains
competitive on match-based metrics even though both models are much weaker on
functional correctness.

The execution metrics give a different view of the task. Execution Accuracy is
higher than Computational Accuracy in all four baseline settings. For example,
the Java-to-Python baselines execute in more than 92\% of the tested cases, but
their Computational Accuracy is only 2.8 for Seq2Seq with attention and 0.4 for
the Transformer. This confirms that program translation is harder than producing
code that simply compiles and runs. A model can generate code that looks correct
and can execute, but the program must also preserve the algorithmic behavior of
the source program. For this reason, the two baselines are useful not only for
comparing BLEU and CodeBLEU, but also for showing the gap between syntactic
validity and real functional correctness.

\section{Zero-Shot Solution with Qwen2.5-Coder}

\quad The first new solution implemented in this project is a zero-shot
translator based on Qwen2.5-Coder-7B-Instruct. Qwen2.5-Coder is a family of
code-oriented large language models trained for code generation, reasoning, and
repair tasks \cite{hui2024qwen25coder}. The selected checkpoint has 7.61
billion parameters \cite{qwen2024qwen25coder7b}. This model was chosen because
it is newer and more code-specialized than the models used in the AVATAR paper,
while still being small enough to run on free Kaggle hardware with 4-bit
quantization.

The method is zero-shot because the model is not fine-tuned on AVATAR and the
prompt contains no examples. For each test program, the script gives the model a
short instruction and the token-spaced source program. The two translation
directions, Java-to-Python and Python-to-Java, are processed in the same run in
round-robin batches of 12 examples. Each completed batch is saved as a separate
prediction file, so the run can be resumed if the Kaggle session stops.

\begin{promptbox}[title={System prompts}]
Java-to-Python:
You translate competitive-programming programs from token-spaced java into valid python. Return only python source code.

Python-to-Java:
You translate competitive-programming programs from token-spaced python into valid java. Return only java source code.
\end{promptbox}

The user prompt asks the model to reconstruct the source program from the
token-spaced input and then write complete normal source code in the target
language. It also contains direction-specific restrictions. For Python output,
the prompt forbids Java-specific constructs such as \texttt{Scanner},
\texttt{System.out.println}, \texttt{charAt}, and \texttt{toCharArray}. For Java
output, it forbids Python-specific constructs such as \texttt{def},
\texttt{range}, \texttt{print}, \texttt{zfill}, and invalid translations of
Python exponentiation such as \texttt{* *}. The decoding is deterministic:
sampling is disabled and the temperature is set to zero.

\begin{promptbox}[title={Base user prompt template}]
Translate this token-spaced {src_lang} program into normal {tgt_lang} source code.

Important:
- Reconstruct the source program before translating it.
- NEW_LINE means a line break. INDENT and DEDENT describe Python-style block nesting in the source.
- Preserve the algorithm, constants, input/output behavior, and edge cases.
- Write complete runnable {tgt_lang} code.
- Do not copy {src_lang} libraries, methods, wrappers, or syntax into the answer.
- Output normal {tgt_lang} source code, not a token stream.
- Keep hardcoded values hardcoded. Only read input when the source program reads input.
- Mentally check the final code for valid {tgt_lang} syntax before returning it.

Target language: {tgt_lang}
{target_examples}

Return only the translated {tgt_lang} source code. Do not explain. Do not use Markdown.

Source tokenized {src_lang} program:
{code}
\end{promptbox}

\begin{promptbox}[title={Python target prompt addition}]
Python output requirements:
- Output Python code only.
- Do not use Java syntax such as public, static, void, class Main, Scanner, String[] args, System.out.println, System.out.printf, charAt, toCharArray, or semicolons as statement endings.
- Use Python input(), sys.stdin, print(), sys.stdout.write(), lists, strings, math, and normal Python functions/classes where appropriate.
- Translate Java string methods to Python: s.charAt(i) becomes s[i], s.length() becomes len(s), toCharArray() becomes the same string or list(s).
- Translate Java Math methods to Python math functions or operators.
- If Java uses geometry/library objects, reimplement the math with Python functions, tuples, or complex numbers. Do not call methods like intersects(), distance(), p1, or p2 on tuples.
- Avoid shadowing Python builtins such as len, str, input, sum, max, min, list, dict, and set.
\end{promptbox}

\begin{promptbox}[title={Java target prompt addition}]
Java output requirements:
- Output Java code only.
- Do not use Python syntax such as def, print(), range(), while True, import sys, list methods, string zfill, or colon-based blocks.
- Use Java Scanner/BufferedReader/StringTokenizer only when the source reads input. Keep hardcoded test values hardcoded.
- Use System.out.print/println/printf, arrays, ArrayList, StringBuilder, Math, braces, and semicolons where appropriate.
- Translate Python ** to Java Math.pow(...) or direct multiplication; never output * *.
- Translate Python zfill(n) to zero padding with String.format(...).replace(' ', '0') or equivalent Java code.
- Initialize boolean sieve arrays correctly, usually with Arrays.fill(prime, true), before marking composites.
- Methods called from main should be static unless they are called through an object.
\end{promptbox}

Early tests showed that asking the model to directly generate the AVATAR token
stream was unreliable. Because of this, the final script asks Qwen to generate
normal target-language source code first. A deterministic post-processing step
then converts this code back into the tokenized format used by AVATAR. For
Python, the tokenizer recovers \texttt{NEW\_LINE}, \texttt{INDENT}, and
\texttt{DEDENT}; for Java, it produces a flat lexical token stream. The
post-processor also keeps important literals intact, for example
\texttt{"01809"} and \texttt{'0'}, while still representing spaces inside
natural-language strings with the dataset placeholder style.

This solution is useful because it tests how far a recent instruction-tuned code
model can go without task-specific training. Its main limitation is that the
model can still produce code that looks plausible but is not functionally
correct. In the first generated batches, some predictions still contained
incorrect imports, source-language idioms, or small algorithmic changes. For
this reason, the zero-shot solution is treated as a modern LLM baseline, not as
a guaranteed correct translator.
