% !TeX root = main.tex
\chapter{Experimental Setup}

\section{Zero-Shot Qwen2.5-Coder}

\quad The zero-shot experiment is implemented in
\texttt{zeroshot\_qwen\_kaggle.py}. It uses
\texttt{Qwen/}\allowbreak\texttt{Qwen2.5-Coder-7B-Instruct} with the Hugging
Face causal language model interface \cite{wolf2020transformers}. The model is
loaded on one GPU with 4-bit NF4 quantization, half-precision compute, remote
model code enabled, and left-padded tokenization. No AVATAR training is
performed. The script evaluates the full test split in both directions,
Java-to-Python and Python-to-Java.

The run configuration disables sampling, uses one decoding beam, sets the
temperature to zero, keeps the repetition penalty at 1.0, and allows up to 1024
new tokens. The batch size is 12 prompts and the seed is 0.

\section{Encoder-Only UniXcoder}

\quad The encoder-only experiment is implemented in
\texttt{encoder\_only\_unixcoder\_kaggle.py}. It uses
\texttt{microsoft/}\allowbreak\texttt{unixcoder-base} \cite{guo2022unixcoder}
as a pretrained source encoder and adds a four-layer Transformer decoder. The
encoder remains trainable, encoder gradient checkpointing is enabled, and the
decoder uses dropout 0.1. The output layer is tied to the encoder token
embedding matrix.

Both directions are trained in the order Python-to-Java and Java-to-Python.
The full train, validation, and test splits are used. Training uses three
epochs, batch size 4, gradient accumulation over 4 steps, maximum source and
target lengths of 512 tokens, label smoothing 0.1, weight decay 0.01, warmup
ratio 0.06, gradient clipping at 1.0, and seed 1234. The encoder learning rate
is \(2 \times 10^{-5}\), while the decoder learning rate is \(1 \times
10^{-4}\).

\section{Decoder-Only Qwen2.5-Coder}

\quad The decoder-only experiment is implemented in
\texttt{decoder\_only\_qwen\_kaggle.py}. It uses
\texttt{Qwen/}\allowbreak\texttt{Qwen2.5-Coder-0.5B-Instruct}
\cite{qwen2024qwen25coder05b}. The model is loaded in 4-bit NF4 precision with
\texttt{float16} compute and fine-tuned with LoRA adapters
\cite{hu2022lora,dettmers2023qlora}. The LoRA rank is 4, \(\alpha=8\), dropout
is 0.05, and the target modules are \texttt{q\_proj} and \texttt{v\_proj}.
Gradient checkpointing is disabled for this run.

The script trains Python-to-Java and Java-to-Python. The run lasts one epoch
with batch size 4, gradient accumulation over 4 steps, learning rate \(3
\times 10^{-4}\), no weight decay, warmup ratio 0.06, gradient clipping at 1.0,
two dataloader workers, prefetch factor 2, and seed 1234. Source and target
lengths are capped at 320 tokens, with a maximum causal sequence length of 640
tokens.

\section{Seq2Seq CodeT5+}

\quad The Seq2Seq experiment is implemented in
\texttt{seq2seq\_codet5p\_kaggle.py}. It uses
\texttt{Salesforce/}\allowbreak\texttt{codet5p-220m}. The model is loaded with
the Hugging Face sequence-to-sequence interface \cite{wolf2020transformers};
the fast tokenizer is preferred, and the script falls back to a local
byte-level BPE tokenizer if the fast tokenizer fails validation.

The task order is Java-to-Python followed by Python-to-Java, and the full
AVATAR splits are used. Training uses two epochs, batch size 4, gradient
accumulation over 4 steps, maximum source and target lengths of 512 tokens,
learning rate \(5 \times 10^{-5}\), AdamW weight decay 0.01, warmup ratio
0.06, gradient clipping at 1.0, automatic mixed precision enabled, gradient
checkpointing disabled, two dataloader workers, and seed 1234.
