% !TeX root = main.tex
\chapter{Results and Discussion}

\section{Reference Results from the AVATAR Paper}

\quad The original AVATAR paper evaluates several neural models on Java-Python
program translation \cite{ahmad2023avatar}. Table
\ref{tab:avatar-paper-reference} reproduces the most useful columns from the
paper's program-translation results: BLEU for token-level similarity, CodeBLEU
for code-aware similarity, and Computational Accuracy for functional
correctness. The full paper table also reports Syntax Match, Dataflow Match,
and Execution Accuracy.

\begin{table}[htbp]
\centering
\caption{Reference results reported in the original AVATAR paper.}
\label{tab:avatar-paper-reference}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lrrrrrr}
\hline
Model & \multicolumn{3}{c}{Java-to-Python} & \multicolumn{3}{c}{Python-to-Java} \\
\cline{2-4}\cline{5-7}
 & BLEU & CB & CA & BLEU & CB & CA \\
\hline
TransCoder & 38.7 & 36.4 & 0.0 & 45.2 & 32.4 & 0.0 \\
DOBF & 42.0 & 38.9 & 0.0 & 42.3 & 31.2 & 0.0 \\
TransCoder-ST (zero-shot) & 41.7 & 39.3 & 0.0 & 42.5 & 30.7 & 0.0 \\
Seq2Seq + Attention & 57.4 & 42.6 & 2.8 & 59.5 & 43.0 & 0.8 \\
Transformer & 39.6 & 34.8 & 0.4 & 43.5 & 35.6 & 0.4 \\
CodeGPT & 46.3 & 30.2 & 2.8 & 48.9 & 38.0 & 2.0 \\
CodeGPT-adapted & 44.3 & 29.3 & 2.4 & 48.0 & 36.7 & 0.8 \\
CodeBERT & 51.1 & 35.0 & 0.4 & 35.1 & 33.2 & 0.0 \\
GraphCodeBERT & 57.9 & 39.0 & 2.0 & 38.3 & 36.9 & 0.0 \\
PLBART & 63.1 & 46.2 & 6.8 & 69.7 & 48.8 & 0.8 \\
CodeT5 & 62.7 & 46.2 & 6.0 & 60.8 & 50.3 & 1.6 \\
TransCoder-ST (fine-tuned) & 55.4 & 43.8 & 5.6 & 66.0 & 48.6 & 2.0 \\
\hline
\end{tabular}%
}
\end{table}

The reference results show why execution-based evaluation is necessary for this
task. Several models obtain BLEU scores above 55 and CodeBLEU scores above 40,
but their Computational Accuracy remains low. The best Java-to-Python value in
the paper's program-translation table is 6.8, obtained by PLBART, while the
best Python-to-Java value is 2.0, obtained by CodeGPT and the fine-tuned
TransCoder-ST system.

\section{Results of the Implemented Models}

\quad Table \ref{tab:implemented-model-results} summarizes the four implemented
model categories. The same compact metric set is used as in Table
\ref{tab:avatar-paper-reference}, so the thesis results can be compared
directly with the reference systems from the AVATAR paper.

\begin{table}[htbp]
\centering
\caption{Results of the implemented models on the AVATAR test split.}
\label{tab:implemented-model-results}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lrrrrrr}
\hline
Model & \multicolumn{3}{c}{Java-to-Python} & \multicolumn{3}{c}{Python-to-Java} \\
\cline{2-4}\cline{5-7}
 & BLEU & CB & CA & BLEU & CB & CA \\
\hline
Zero-shot Qwen2.5-Coder 7B & 58.23 & 45.67 & 59.68 & 61.02 & 48.51 & 61.80 \\
Encoder-only UniXcoder & 55.20 & 44.20 & 8.00 & 49.44 & 38.18 & 1.95 \\
Decoder-only Qwen2.5-Coder 0.5B & 55.58 & 36.49 & 24.68 & 46.59 & 37.69 & 11.34 \\
Seq2Seq CodeT5+ 220M & 73.11 & 47.37 & 47.82 & 72.07 & 51.76 & 41.29 \\
\hline
\end{tabular}%
}
\end{table}

The strongest implemented model by Computational Accuracy is the 7B zero-shot
Qwen system. It reaches 59.68 for Java-to-Python and 61.80 for Python-to-Java
without AVATAR-specific training. Compared with the AVATAR paper reference
results, it is far above every program-translation model in Table
\ref{tab:avatar-paper-reference} on functional correctness. Its BLEU and
CodeBLEU are not the highest implemented values, which shows that surface
similarity is not the only signal that matters for executable translation.

Seq2Seq CodeT5+ 220M is the strongest implemented fine-tuned model. It obtains
the highest BLEU scores among the implemented systems, 73.11 and 72.07, and the
highest implemented CodeBLEU score for Python-to-Java, 51.76. Its
Computational Accuracy is also strong, with 47.82 for Java-to-Python and 41.29
for Python-to-Java. This places it well above the CodeT5 and PLBART reference
results from the AVATAR paper on functional correctness, although still below
the zero-shot Qwen system used in this thesis.

The decoder-only Qwen2.5-Coder 0.5B model sits between the strong fine-tuned
encoder-decoder model and the weaker encoder-only experiment. Its
Computational Accuracy values, 24.68 and 11.34, are above the reference range
reported in the AVATAR paper's program-translation table, but they do not reach
the CodeT5+ results. This suggests that the causal model can learn useful
translation behavior after fine-tuning, but the smaller checkpoint and the
single-stream prompt format limit functional correctness.

The encoder-only UniXcoder model has reasonable lexical and CodeBLEU scores,
but weak functional results. In Java-to-Python, it reaches 55.20 BLEU and 44.20
CodeBLEU, close to several reference models from the AVATAR paper, yet its
Computational Accuracy is only 8.00. In Python-to-Java, the gap is larger:
49.44 BLEU and 38.18 CodeBLEU correspond to only 1.95 Computational Accuracy.
This is near the top of the original paper's Python-to-Java functional range,
but it remains weak in absolute terms. The result confirms that a pretrained
source encoder is not enough by itself; the target-side generator is critical
for producing behaviorally correct programs.

\section{Discussion}

\quad The main pattern across both tables is the separation between textual
similarity and functional correctness. BLEU and CodeBLEU reward programs that
look close to the reference, but AVATAR requires preserving runtime behavior.
This distinction explains why models such as PLBART, CodeT5, and
GraphCodeBERT can have respectable BLEU and CodeBLEU values while still
producing low Computational Accuracy in the original paper.

The implemented results follow the same pattern. CodeT5+ is the best model for
surface similarity, but zero-shot Qwen2.5-Coder 7B is the best model for
functional correctness. This difference is important: the larger instruction
model does not reproduce the reference format as closely as the fine-tuned
Seq2Seq model, but it more often preserves the behavior of the source program.
For a program translation system, this is the more important property.

The comparison with the AVATAR paper also shows that model scale and pretraining
objective matter. The zero-shot Qwen model exceeds the reference systems on
functional correctness even without supervised training on the benchmark, while
the smaller decoder-only model improves over the original functional range but
does not reach the same level. The encoder-only model gives the clearest
negative result: reusing only a pretrained encoder helps produce plausible text,
but it does not solve the harder problem of generating correct target-language
programs.

Overall, the best result of this thesis is not the model with the highest BLEU,
but the model with the highest Computational Accuracy. The results therefore
support the central evaluation choice of the project: code translation should
be judged by behavioral correctness, with lexical and structural metrics used
as supporting evidence rather than as the final criterion.
